---
title: "Data Visualization for Environmental Epidemiology with ggplot2: Mastering Presentation Grade Figures"
author: "Alexandra E. Larsen, Alison K. Krajewski, Lauren H. Wyatt"
date: "7/7/2020"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Dependencies

These are the libraries and packages we will be using today. If any of them are not loaded already in your RStudio, use [install.packages("library-name")].

```{r, include = T, echo = T, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyr)
library(tidyverse)
library(datasets)
library(RColorBrewer)
library(colorRamps)
library(sf)
library(maps)
library(dplyr)
library(data.table)
library(stringr)
library(viridis)
library(psych)
library(mlbench)
library(extrafont)
```

# Datasets

```{r}
# Built-in data sets
data(iris)
data(faithful)
data(cars)
data(mpg)
data(Ozone)

# Created by Alison and Lauren
setwd("C:\\Users\\lwyatt\\Desktop\\Wyatt\\Workshops\\DataViz_ggplot2\\Data\\Data_for_github\\") # update with data location folder
air          <- read.csv("Criteria_Air_Pollutants.csv")
ht_top5      <-  read.csv("Hypertension_Top5.csv")
ht_top5_wide <- read.csv("Hypertension_Top5_Wide.csv")
air_long     <- read.csv("CriteriaAirPollutants_Long.csv")
```

# Formating Data for GGPlot2

There are many libraries available in R for data manipulation, but we will be focusing on the package [tidyr]. Tidyr uses a pipe operator, just like ggplot2. We've loaded tidyr in the Dependencies section. 

## The Pipe Operator

Let's start with some exercises to get familiar with the pipe operator, [%>%]. For this example, we'll generate some N(0,1) data and take the mean.

```{r}
set.seed(1234)
sample_data <- rnorm(n = 1000)
mean(sample_data)
```

We can use a pipe operator like so:

```{r}
sample_data %>%
  mean()
```

Here is another example where we create a barplot of some of the iris data.

```{r}
head(iris)
barplot(colMeans(iris[iris$Species == "setosa", 1:4]), main = "Barplot without Pipe Operator")
```

Using the pipe operator instead:

```{r}
iris %>%
  filter(Species == "setosa") %>%
  select(Sepal.Length:Petal.Width) %>%
  colMeans() %>%
  barplot(main = "Barplot with Pipe Operator")
```

In the example using the pipe operator above, notice that we didn't have to create any variables and we can avoid long lines of code. Futher, tidyr has intuitive functions that we can take advantage of as we'll see in the next section.

## Long vs. Wide

One of the most common data transformations that you need with using ggplot2() is converting long to wide and visa versa. 

Tidyr refers to changing from long to wide as "pivoting" and uses pivot_longer() and pivot_wider(). The pivot functions have replaced gather() and spread() as the tidyr function for lengthening or widening data. Other packages are available to do this - you may be familiar with Reshape2, which uses melt() and dcast().

A good resource on gather(), spread(), or the reshape2 options is [here](http://www.cookbook-r.com/Manipulating_data/Converting_data_between_wide_and_long_format/).

We'll start with going from *wide to long data* in the iris data. 

```{r}
iris_long <- iris %>%  
  pivot_longer(
    cols = Sepal.Length:Petal.Width,
    names_to = "Part",
    values_to = "Measurement")
head(iris_long)
```

```{r}
# going back to wide... need another data set?

```

# Basics of ggplot2

In this section, we'll spend time with data and aesthetic mapping, and try out a few types of graphs.

Let's start with a basic scatterplot of eruptions from Old Faithful.

```{r}
head(faithful)
```

```{r}
ggplot(data = faithful, 
       mapping = aes(x = waiting, y = eruptions)) +
  geom_point()
```

We can specify data and aesthetics in the layer too

```{r}
ggplot() +
  geom_point(data = faithful, 
             mapping = aes(x = waiting, y = eruptions))
```

Or a mixture of both
```{r}
ggplot(data = faithful) +
  geom_point(mapping = aes(x = waiting, y = eruptions))
```

And the data doesn't have to come first.
```{r}
ggplot(mapping = aes(x = waiting, y = eruptions)) + 
  geom_point(data = faithful)
```

But, you do need both. If you don't specify data within the ggplot, you need $:

```{r eval = F}
ggplot(mapping = aes(x = waiting, y = eruptions)) + 
  geom_point()
```

And if you don't specify mapping, nothing will happen. Remember, mapping is what connects the data to your plot.

```{r, eval = F}
ggplot(data = faithful) +
  geom_point()
```

Without the mapping argument and aes, can't "talk" to the variables in the data.

```{r eval = F}
ggplot() + 
  geom_point(data = faithful, x = waiting, y = eruptions)
```

If you want to change an aesthetic, say color, based on the data, include in mapping

```{r}
ggplot() + 
  geom_point(data = faithful, 
             mapping = aes(x = waiting, y = eruptions, color = eruptions > 3))
```

Otherwise, place it outside of mapping; it will apply to all the data.

```{r}
ggplot() + 
  geom_point(data = faithful, 
             mapping = aes(x = waiting, y = eruptions),
             color = "blue")
```

Let's look at a histogram of the eruptions. 

```{r}
ggplot() +
  geom_histogram(data = faithful,
                 mapping = aes(x = eruptions))
```

Change with width of each bin from default (30)

```{r}
ggplot() +
  geom_histogram(data = faithful,
                 mapping = aes(x = eruptions),
                 bins = 50)
```

Let's make a box plot with the setosa data from iris.

```{r}
setosa <- iris %>%
  filter(Species == "setosa") %>%
  select(-Species)
head(setosa)
```

For the boxplot, we'll want the parts of the flower on the x-axis and the measurements on the y-axis, so we need "parts" and "measurements" to be variables or columns in the data set. To get there, we can pivot the data to long format like we did earlier. Let's re-use that code for our setosa subset:

```{r}
setosa_long <- setosa %>%
  pivot_longer(
    cols = Sepal.Length:Petal.Width,
    names_to = "Part",
    values_to = "Measurement")
head(setosa_long)
```

Now, we can make our boxplot.

```{r}
ggplot() +
  geom_boxplot(data = setosa_long, aes(x = Part, y = Measurement))
```

# Scales

Everything inside the aes() will have scales

## Modify Scales

Scales are made up of three pieces separated by an underscore (_)

Without specifying scales, the defaults are used. 
```{r}
#Example of Basic Sccatterplot Without Specifying Scales
ozone_plot <- ggplot(air, aes(Year, Ozone_ppm)) + geom_point()
ozone_plot
```

## Breaks and Labels

This example changes the lab name from the variable using scales.

```{r}
#Adding in Labels 
ozone_plot <- ggplot(air, aes(x=Year, y=Ozone_ppm)) + geom_point()
ozone_plot + scale_x_continuous(name="Year") + scale_y_continuous(name="Ozone Concentrations (ppm)")
```

This examples uses the xlab and ylab options.

```{r}
#Another way to add labels
ozone_plot <- ggplot(air, aes(x=Year, y=Ozone_ppm)) + geom_point() +
  xlab("Year") + ylab("Ozone Concentrations (ppm)")
ozone_plot 
```

Labels can also be created for mathematical expressions.

```{r}
#Example of X-Axis Title with A Mathematicel Expression to the Title
ozone_plot <- ggplot(air, aes(x=Year, y=Ozone_ppm)) + geom_point()
ozone_plot + scale_x_continuous(name="Year") + 
  scale_y_continuous(quote(Ozone + Concentrations ^ ppm))
```

```{r}
#Another example with Superscripts and Subscripts
NO2_plot_script <- ggplot(air, aes(x=Year, y=Nitrogen_Dioxide_ppm)) + geom_point() +
  xlab("Year") + ylab(expression(Nitrogen~Dioxide~(NO[2])~(ppm)))
NO2_plot_script
```

In addition to modifying the scales, you can also modify limits.

```{r}
#Example with Modifying Limits
ozone_plot <- ggplot(air, aes(x=Year, y=Ozone_ppm)) + geom_point(aes(colour=Year)) +
  scale_x_continuous(name="Year", limits=c(2000,2020)) + 
  scale_y_continuous(name="Ozone Concentrations (ppm)", limits=c(0.050, 0.090)) 
ozone_plot
```

Break points at specified points can also be created.

```{r}
#Example of Breaks
#Ozone Example
ozone_plot <- ggplot(air, aes(x=Year, y=Ozone_ppm)) + geom_point(aes(colour=Year)) +
  scale_x_continuous(name="Year", limits=c(2000,2020)) + 
  scale_y_continuous(name="Ozone Concentrations (ppm)", breaks=c(0.06, 0.07, 0.08, 0.09)) 
ozone_plot
```

Hypertension example with the defaults

```{r}
#Example of Labels
#Hypertension Example - Default
ht_top5_plot_labels <- ggplot(ht_top5, aes(x=State, y=Total, group=Sex, colour=sex)) + 
  geom_point(aes(colour=Sex), size=5) +
  scale_x_discrete(name="State") +
  scale_y_discrete(name="Total Hypertension Prevelance (%)")
ht_top5_plot_labels
```

Hypertensions with breaks and labels

```{r}
#Hypertension Example with Break and Labels
ht_top5_plot_labels <- ggplot(ht_top5, aes(x=State, y=Total, group=Sex, colour=sex)) + 
  geom_point(aes(colour=Sex), size=5) +
  scale_x_discrete(name="State", labels=c("Alabama" = "AL", "District Of Columbia" = "DC", "Mississippi" = "MS", 
                            "South Carolina" = "SC", "West Virginia" = "WV")) +
  scale_y_discrete(name="Total Hypertension Prevelance (%)")
ht_top5_plot_labels
```

## Legends

### Changing the Position of the Legend

The default position is to the right of the figure.

```{r}
ozone_plot <- ggplot(air, aes(x=Year, y=Ozone_ppm)) + geom_point(aes(colour=Year)) +
  scale_x_continuous(name="Year", limits=c(2000,2020)) + 
  scale_y_continuous(name="Ozone Concentrations (ppm)") 
ozone_plot
```

The designation of the legend to the right of the figure can also be specified.

```{r}
ozone_plot <- ggplot(air, aes(x=Year, y=Ozone_ppm)) + geom_point(aes(colour=Year)) +
  scale_x_continuous(name="Year", limits=c(2000,2020)) + 
  scale_y_continuous(name="Ozone Concentrations (ppm)") +
  theme(legend.position='right')
ozone_plot
```

Positioning the legend on the bottom.

```{r}
#Ozone Example
ozone_plot <- ggplot(air, aes(x=Year, y=Ozone_ppm)) + geom_point(aes(colour=Year)) +
  scale_x_continuous(name="Year", limits=c(2000,2020)) + 
  scale_y_continuous(name="Ozone Concentrations (ppm)") +
  theme(legend.position="bottom")
ozone_plot
```

The legend can also be modified so that all the values of the legend will be in single row.

```{r}
#Hypertension Example
ht_top5_plot_legend <- ggplot(ht_top5, aes(x=State, y=Total, group=Sex, colour=sex)) + 
  geom_point(aes(colour=Sex), size=5) +
  scale_x_discrete(name="State", labels=c("Alabama" = "AL", "District Of Columbia" = "DC", "Mississippi" = "MS", 
                            "South Carolina" = "SC", "West Virginia" = "WV")) +
  scale_y_discrete(name="Total Hypertension Prevelance (%)") + 
  theme(legend.position="bottom") + guides(colour=guide_legend(nrow=1))
ht_top5_plot_legend
```

The legend can also be removed.

```{r}
ozone_plot <- ggplot(air, aes(x=Year, y=Ozone_ppm)) + geom_point(aes(colour=Year)) +
  scale_x_continuous(name="Year", limits=c(2000,2020)) + 
  scale_y_continuous(name="Ozone Concentrations (ppm)") +
  theme(legend.position="none")
ozone_plot
```

## Colors

General considerations:
* Avoid using red-green gradients
* Red is alarming and can imply something bad
* Avoid using rainbow gradients
* Keep color schemes color blind friendly (Reference: http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette)

```{r}
#Adding in a scale for color
#Ozone Example
ozone_plot <- ggplot(air, aes(x=Year, y=Ozone_ppm)) + geom_point(aes(colour=Year), size=5) +
  scale_x_continuous(name="Year") + 
  scale_y_continuous(name="Ozone Concentrations (ppm)") 
ozone_plot
```

Hypertension examples with default colors

```{r}
#Hypertension Example
ht_top5_plot <- ggplot(ht_top5, aes(x=State, y=Total)) + geom_point(aes(colour=Sex), size=5) +
  scale_x_discrete(name="State") +
  scale_y_discrete(name="Total Hypertension Prevelance (%)")
ht_top5_plot

```

Using the hypertension example to add in a specificed color option

```{r}
#Example using scale_color_brewer
ht_top5_plot <- ggplot(ht_top5, aes(x=State, y=Total)) + geom_point(aes(colour=Sex), size=5) +
  scale_x_discrete(name="State") +
  scale_y_discrete(name="Total Hypertension Prevelance (%)") +
  scale_color_brewer()
ht_top5_plot
```

Changing the color manually

```{r}
#Example using scale_color_brewer
ht_top5_plot_color <- ggplot(ht_top5, aes(x=State, y=Total)) + geom_point(aes(colour=Sex), size=5) +
  scale_x_discrete(name="State") +
  scale_y_discrete(name="Total Hypertension Prevelance (%)") +
  scale_color_brewer(palette = "Greens")
ht_top5_plot_color
```
```{r}
#Example using scale_color_brewer
ht_top5_plot_color <- ggplot(ht_top5, aes(x=State, y=Total)) + geom_point(aes(colour=Sex), size=5) +
  scale_x_discrete(name="State") +
  scale_y_discrete(name="Total Hypertension Prevelance (%)") +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position="bottom") + guides(colour=guide_legend(nrow=1))
ht_top5_plot_color
```
\newpage

# Themes

```{r multiplot definition 1, echo=FALSE}
external_data_dir <- "C:\\Users\\lwyatt\\Desktop\\Wyatt\\Workshops\\DataViz_ggplot2\\Data\\Data_for_github\\"
```

## What is a theme?

Themes allow for the customization of the non-data parts of figures (ie. titles, labels, fonts, background, gridlines, legends).

## Pre-existing themes

```{r multiplot definition, echo=FALSE}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```


```{r theme defaults, message=FALSE, warning=FALSE, fig.width=6, fig.height=8}
base_plot <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(show.legend = FALSE) +
  scale_color_brewer(palette = "Dark2")

default_plot_title <- base_plot +
  labs(title = "Default (theme_grey)")

bw_plot_title <- base_plot +
  labs(title = "theme_bw") +
  theme_bw()

linedraw_plot_title <- base_plot +
  labs(title = "theme_linedraw") +
  theme_linedraw()

light_plot_title <- base_plot +
  labs(title = "theme_light") +
  theme_light()

dark_plot_title <- base_plot +
  labs(title = "theme_dark") +
  theme_dark()

minimal_plot_title <- base_plot +
  labs(title = "theme_minimal") +
  theme_minimal()

classic_plot_title <- base_plot +
  labs(title = "theme_classic") +
  theme_classic()

void_plot_title <- base_plot +
  labs(title = "theme_void") +
  theme_void()

multiplot(default_plot_title, bw_plot_title, linedraw_plot_title,
          light_plot_title, dark_plot_title, minimal_plot_title,
          classic_plot_title, void_plot_title, layout = matrix(c(1,2,3,4,5,6,7,8), 
                                                               nrow = 4, byrow = TRUE))
```

```{r theme improve scatter save, message=FALSE, warning=FALSE, eval=FALSE, echo=FALSE}

# jpeg("C:\\Users\\lwyatt\\Desktop\\Wyatt\\Workshops\\DataViz_ggplot2\\Theme_pre_ex1.jpeg", 
#      units = "in", width = 8, height = 10, res = 400)
# multiplot(default_plot_title, bw_plot_title, linedraw_plot_title,
#           light_plot_title, dark_plot_title, minimal_plot_title,
#           classic_plot_title, void_plot_title, layout = matrix(c(1,2,3,4,5,6,7,8), nrow = 4, byrow = TRUE))
# dev.off()
# 
# jpeg("C:\\Users\\lwyatt\\Desktop\\Wyatt\\Workshops\\DataViz_ggplot2\\Theme_pre_ex2.jpeg", 
#      units = "in", width = 6, height = 8, res = 400)
# multiplot(default_plot_title, bw_plot_title, linedraw_plot_title,
#           light_plot_title, dark_plot_title, minimal_plot_title,
#           classic_plot_title, void_plot_title, layout = matrix(c(1,2,3,4,5,6,7,8), nrow = 4, byrow = TRUE))
# dev.off()

```
\newpage

## Scatterplot example

```{r theme improve scatter data create, message=FALSE, warning=FALSE, fig.width=8, fig.height=3, eval=FALSE, echo=FALSE}
####################################################
#
### Data prep occured prior to workshop
#
####################################################

### Load Los Angeles Ozone Pollution Data, 1976 (package: mlbench)
# subsititute temp data to create ozone data for a second city
Ozone_sub <- Ozone[, c(1,2,3,4,8)]
colnames(Ozone_sub) <- c("Month", "Day", "DOW", "City A", "City B")
Ozone_sub$Date <- as.Date(paste(Ozone_sub$Month, Ozone_sub$Day, "1976", sep = "-"), format = "%m-%d-%Y")
Ozone_sub$`City B` <- Ozone_sub$`City B` / 2
head(Ozone_sub)

Ozone_sub_long <- gather(Ozone_sub, Env_var, measurement, `City A`:`City B`)
head(Ozone_sub_long)

saveRDS(Ozone_sub_long, "C:\\Users\\lwyatt\\Desktop\\Wyatt\\Workshops\\DataViz_ggplot2\\Data\\Data_for_github\\Ozone_ex\\Ozone_ex.rds")
```

Original default plot

```{r theme improve scatter data load, message=FALSE, warning=FALSE, fig.width=8, fig.height=3}
### Load altered Los Angeles Ozone Pollution Data, 1976 (package: mlbench)
# ozone data for a second city was created by subsitituting temp data 

Ozone_sub_long <- readRDS(paste0(external_data_dir, "Ozone_ex\\Ozone_ex.rds"))

base_plot <- ggplot(Ozone_sub_long, aes(x = Date, y = measurement, color = Env_var)) +
  geom_point() +
  scale_color_brewer(palette = "Dark2") +
  facet_grid(. ~ Env_var) +
  labs(x = "",
       y = "Daily maximum \n1-hr average ozone")
base_plot
```
\newpage

Originial plot with only theme_bw

```{r theme improve scatter 1, message=FALSE, warning=FALSE, fig.width=8, fig.height=3}
bw_plot <- base_plot +
  theme_bw()
bw_plot
```

Adjustment to reduce gridlines

```{r theme improve scatter 2, message=FALSE, warning=FALSE, fig.width=8, fig.height=3}
bw_plot_adj1 <- base_plot +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank()
  )
bw_plot_adj1
```
\newpage

Adjustment to increase global text size

```{r theme improve scatter 3, message=FALSE, warning=FALSE, fig.width=8, fig.height=3}
text_size <- 15

bw_plot_adj2 <- base_plot +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    text = element_text(size = text_size)
)
bw_plot_adj2
```

Adjustment to increase facet label text size

```{r theme improve scatter 4, message=FALSE, warning=FALSE, fig.width=8, fig.height=3}
bw_plot_adj3 <- base_plot +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    text = element_text(size = text_size),
    strip.text.x = element_text(size = 20)
)
bw_plot_adj3
```
\newpage

Adjustment to rotate x-axis labels

```{r theme improve scatter 5, message=FALSE, warning=FALSE, fig.width=8, fig.height=3}
bw_plot_adj4 <- base_plot +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    text = element_text(size = text_size),
    strip.text.x = element_text(size = 20),
    axis.text.x = element_text(angle = 45, hjust = 1)
)
bw_plot_adj4
```

Adjustment to change spacing between facet panels

```{r theme improve scatter 6, message=FALSE, warning=FALSE, fig.width=8, fig.height=3}
bw_plot_adj5 <- base_plot +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    text = element_text(size = text_size),
    strip.text.x = element_text(size = 20),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.spacing = unit(2, "lines")
)
bw_plot_adj5
```
\newpage

Adjustment to change legend (position, text size)

```{r theme improve scatter 7, message=FALSE, warning=FALSE, fig.width=8, fig.height=4}
bw_plot_adj6 <- base_plot +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    text = element_text(size = text_size),
    strip.text.x = element_text(size = 20),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.spacing = unit(2, "lines"),
    legend.position = "bottom"
)
bw_plot_adj6
```


```{r theme improve scatter 8, message=FALSE, warning=FALSE, fig.width=8, fig.height=3, eval=FALSE, echo=FALSE}
# Adjustment to change font (extrafont library)
font_import()
loadfonts(device = "win") # long process

bw_plot_adj7 <- base_plot +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    text = element_text(size = text_size, family = "Trebuchet MS"),
    strip.text.x = element_text(size = 20),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.spacing = unit(2, "lines"),
    legend.position = "bottom"
)
bw_plot_adj7
```
\newpage


# Facets

Facets generate small groupings with a different subset of the data.

##Facet Wrap
Using the hypertension exmaple, two individual plots by sex are created.

Males:
```{r}
#Hypertension Example of Fracet Wrapping
#Hypertension Prevalence by Sex
#Males
ht_male <- ggplot(ht_top5_wide, aes(x=State, y=Total_Male)) + geom_point(size=5, color="darkblue") +
  scale_x_discrete(name="State") + 
  scale_y_discrete(name="Total Hypertension Prevalence for Males (%)") +
  theme_bw()
ht_male
```
Females:
```{r}
#Females
ht_female <- ggplot(ht_top5_wide, aes(x=State, y=Total_Female)) + geom_point(size=5, color="deeppink") +
  scale_x_discrete(name="State") + 
  scale_y_discrete(name="Total Hypertension Prevalence for Females (%)") +
  theme_bw()
ht_female
```
Using facet wrap to combine the two together.
```{r}
#Using Facet Wrap
ht_top5_plot <- ggplot(ht_top5, aes(x=State, y=Total, group=Sex, colour=Sex)) + geom_point(size=5) +
  scale_x_discrete(name="State", labels=c("Alabama" = "AL", "District Of Columbia" = "DC", "Mississippi" = "MS", 
                            "South Carolina" = "SC", "West Virginia" = "WV")) + 
  scale_y_discrete(name="Total Hypertension Prevalence (%)") +
  theme_bw() + 
  theme(legend.position='bottom') + guides(colour = guide_legend(nrow = 1)) +
  facet_wrap(~Sex, scales = "free")
ht_top5_plot
```

```{r}
#This produces the same graph as above, but slightly different coding for facet_wrap
ht_top5_plot <- ggplot(ht_top5, aes(x=State, y=Total, group=Sex, colour=Sex)) + geom_point(size=5) +
  scale_x_discrete(name="State", labels=c("Alabama" = "AL", "District Of Columbia" = "DC", "Mississippi" = "MS", 
                            "South Carolina" = "SC", "West Virginia" = "WV")) + 
  scale_y_discrete(name="Total Hypertension Prevalence (%)") +
  theme_bw() + 
  theme(legend.position='bottom') + guides(colour = guide_legend(nrow = 1)) +
  facet_wrap(~Sex, nrow = 1, dir="h")
ht_top5_plot

```

## Facet Grid
Creating a vertical grid
```{r}
#Using Facet Grid 
#Vertical
ht_top5_plot_vertical <- ggplot(ht_top5, aes(x=State, y=Total, group=Sex, colour=Sex)) + geom_point(size=5) +
  scale_x_discrete(name="State", labels=c("Alabama" = "AL", "District Of Columbia" = "DC", "Mississippi" = "MS", 
                            "South Carolina" = "SC", "West Virginia" = "WV")) + 
  scale_y_discrete(name="Total Hypertension Prevalence (%)") +
  theme_bw() + 
  theme(legend.position='bottom') + guides(colour = guide_legend(nrow = 1)) +
  facet_grid(~Sex, scales = "free")
ht_top5_plot_vertical
```
Creating a horizontal grid
```{r}
#Horizontal
ht_top5_plot_horizontal <- ggplot(ht_top5, aes(x=State, y=Total, group=Sex, colour=Sex)) + geom_point(size=5) +
  scale_x_discrete(name="State", labels=c("Alabama" = "AL", "District Of Columbia" = "DC", "Mississippi" = "MS", 
                            "South Carolina" = "SC", "West Virginia" = "WV")) + 
  scale_y_discrete(name="Total Hypertension Prevalence (%)") +
  theme_bw() + 
  theme(legend.position='bottom') + guides(colour = guide_legend(nrow = 1)) +
  facet_grid(Sex~., scales = "free")
ht_top5_plot_horizontal
```
Creating a facet with multiple air pollutants
```{r}
#Creating multiple plots for the criteria air pollutants
#Only faceting the plots with the same scale (units)
cap_plot <- ggplot(air_long, aes(x=Year, y=Concentrations, group=Pollutant, color=Pollutant)) + 
  geom_point(size=5) + 
  geom_errorbar(aes(ymin=Lower_10th_Percentile, ymax=Upper_90th_Percentile), width=0.5, size=2) + 
  scale_x_discrete(name="Year") + 
  scale_y_continuous(name="Concentrations (ppm)") +
  theme_bw() +
  facet_grid(Pollutant~., scales = "free")
cap_plot
```
\newpage

# Mapping

## Map example 1 - Fires

```{r map fire data aggregate, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE, echo=FALSE}

####################################################
#
### Data prep occured prior to workshop
#
####################################################

### Create summary dataset for choropleth map example (US counties, contiguous US)
### Data will be filtered to the years 2008 and 2009 and aggregated to the number of fires reported per county
#      Source: U.S. Department of Agriculture, https://www.fs.usda.gov/rds/archive/Catalog/RDS-2013-0009
#      Source title: Spatial wildfire occurrence data for the United States, 1992-2011
#      Notes: The uploaded file (FPA_FOD_20130422.csv) contains the first 65534 rows from the wildfire database

# fire_dir <- "...\\Fire_ex\\"
fires <- fread(paste0(external_data_dir, "Fire_ex\\FPA_FOD_20130422.csv"))
fires <- fires[,c("FPA_ID", "FIRE_YEAR", "STATE", "FIPS_CODE", "DISCOVERY_DATE", "CONTAIN_DATE", "STAT_CAUSE_DESCR")]

# crosswalk between state character and numeric codes
st_FIPS_xwalk <- fread(paste0(external_data_dir, "Fire_ex\\State_FIPS_crosswalk.csv"))
st_FIPS_xwalk$State <- as.factor(str_pad(st_FIPS_xwalk$FIPS, 2, pad = "0"))
st_FIPS_xwalk <- st_FIPS_xwalk[,c("Postal Code", "State")]
colnames(st_FIPS_xwalk) <- c("STATE", "State")

fires <- st_FIPS_xwalk[fires, on = "STATE"]
fires$FIPS <- paste0(fires$State, str_pad(fires$FIPS_CODE, 3, pad = "0"))
fires <- fires[,c("FPA_ID", "FIRE_YEAR", "FIPS", "DISCOVERY_DATE", "CONTAIN_DATE", "STAT_CAUSE_DESCR")]

fire_county_2008_2009 <- fires[fires$FIRE_YEAR %in% c(2008,2009)] %>%
  group_by(FIPS, FIRE_YEAR) %>%
  summarise(fire_sum_total = n(),
            fire_sum_misc = sum(STAT_CAUSE_DESCR == "Miscellaneous"),
            fire_sum_lightning = sum(STAT_CAUSE_DESCR == "Lightning"),
            fire_sum_debris = sum(STAT_CAUSE_DESCR == "Debris Burning"),
            fire_sum_camp = sum(STAT_CAUSE_DESCR == "Campfire"),
            fire_sum_equip = sum(STAT_CAUSE_DESCR == "Equipment Use"),
            fire_sum_arson = sum(STAT_CAUSE_DESCR == "Arson"),
            fire_sum_child = sum(STAT_CAUSE_DESCR == "Children"),
            fire_sum_railroad = sum(STAT_CAUSE_DESCR == "Railroad"),
            fire_sum_smoking = sum(STAT_CAUSE_DESCR == "Smoking")) %>%
  data.table()

saveRDS(fire_county_2008_2009, "C:\\Users\\lwyatt\\Desktop\\Wyatt\\Workshops\\DataViz_ggplot2\\Data\\RDS-2013-0009\\Data\\Fire_ex.rds")
```


```{r map improve map data load, message=FALSE, warning=FALSE, echo=FALSE, results="hide"}
###
#### Data prep
###

### State and county shapefiles (US)
census_shapfile_dir <- paste0(external_data_dir, "Shapefiles\\Census_2010\\")
USstates <- st_read(paste0(census_shapfile_dir, "States2010_outline\\gz_2010_us_040_00_5m\\gz_2010_us_040_00_5m.shp"))
UScounties <- st_read(paste0(census_shapfile_dir, "Counties2010_outline\\gz_2010_us_050_00_5m\\gz_2010_us_050_00_5m.shp"))
UScounties$FIPS <- paste0(UScounties$STATE, UScounties$COUNTY)
yr_list <- data.frame(  FIRE_YEAR = c(2008, 2009) )
UScounties <- merge(UScounties, yr_list) 

### Fire data, aggregated to counts per county in 2008 and 2009
fire_county_2008_2009 <- readRDS(paste0(external_data_dir, "Fire_ex\\Fire_ex.rds"))
fire_county_2008_2009 <- fire_county_2008_2009[,c("FIPS", "FIRE_YEAR", "fire_sum_total")]

### Join aggregated fire data to county shapefile
UScounties <- UScounties %>%
  left_join(fire_county_2008_2009, by = c("FIPS", "FIRE_YEAR"))
UScounties[is.na(UScounties)] <- 0

### Create breaks of fireS per county
break_list <- c(0,1,5,10,20,30,40,50,60, 800)
UScounties$fire_sum_total_cat <- cut(as.numeric(UScounties$fire_sum_total), 
                                     breaks = break_list, labels = c("0", "1-4", "5-9", "10-19", "20-29", "30-39",
                                                                     "40-49", "50-59", "60+"), right = F)
# table(UScounties$fire_sum_total_cat)

### Subset to continuous US (exclude HI, AK, PR)
exclude_HI_AK_PR <- c("02","15","72")
USstates <- USstates %>% filter( !(STATE %in% exclude_HI_AK_PR) )
UScounties <- UScounties %>% filter( !(STATE %in% exclude_HI_AK_PR))
```

Default map for 2008

```{r map improve map 1, message=FALSE, warning=FALSE}
map_1 <- ggplot(subset(UScounties, FIRE_YEAR == 2008), aes(fill = fire_sum_total_cat)) +
  geom_sf() +
  geom_sf(data = USstates, fill = NA) +
  labs(fill = "# of fires")
map_1
```
\newpage

Adjust focus to map, minimizing background, adding conic projection

```{r map improve map 2, message=FALSE, warning=FALSE}
map_2 <- ggplot(subset(UScounties, FIRE_YEAR == 2008), aes(fill = fire_sum_total_cat)) +
  geom_sf() +
  geom_sf(data = USstates, fill = NA) +
  coord_sf(crs = st_crs(5070)) +
  theme_void() +
  labs(fill = "# of fires")
map_2
```
\newpage

Specify color palette

```{r map improve map 3, message=FALSE, warning=FALSE}
map_3 <- ggplot(subset(UScounties, FIRE_YEAR == 2008), aes(fill = fire_sum_total_cat)) +
  geom_sf() +
  geom_sf(data = USstates, fill = NA) +
  scale_fill_brewer(palette = "OrRd") +
  coord_sf(crs = st_crs(5070)) +
  theme_void() +
  labs(fill = "# of fires")
map_3
```
\newpage

Adjust legend so higher values are on top

```{r map improve map 4, message=FALSE, warning=FALSE}
map_4 <- ggplot(subset(UScounties, FIRE_YEAR == 2008), aes(fill = fire_sum_total_cat)) +
  geom_sf() +
  geom_sf(data = USstates, fill = NA) +
  scale_fill_brewer(palette = "OrRd", guide = guide_legend(reverse = TRUE)) +
  coord_sf(crs = st_crs(5070)) +
  theme_void() +
  labs(fill = "# of fires")
map_4
```
\newpage

Change county line color so state shapes are apparent

```{r map improve map 5, message=FALSE, warning=FALSE}
map_5 <- ggplot(subset(UScounties, FIRE_YEAR == 2008), aes(fill = fire_sum_total_cat)) +
  geom_sf(color = "#abb2b9") +
  geom_sf(data = USstates, fill = NA) +
  scale_fill_brewer(palette = "OrRd", guide = guide_legend(reverse = TRUE)) +
  coord_sf(crs = st_crs(5070)) +
  theme_void() +
  labs(fill = "# of fires")
map_5
```
\newpage

Reduce size of county lines

```{r map improve map 6, message=FALSE, warning=FALSE}
cty_line <- 0.2

map_6 <- ggplot(subset(UScounties, FIRE_YEAR == 2008), aes(fill = fire_sum_total_cat)) +
  geom_sf(color = "#abb2b9", size = cty_line) +
  geom_sf(data = USstates, fill = NA) +
  scale_fill_brewer(palette = "OrRd", guide = guide_legend(reverse = TRUE)) +
  coord_sf(crs = st_crs(5070)) +
  theme_void() +
  labs(fill = "# of fires")
map_6
```
\newpage

Facet by year, legend applies to all maps

```{r map improve map 7, message=FALSE, warning=FALSE}
map_7 <- ggplot(UScounties, aes(fill = fire_sum_total_cat)) +
  geom_sf(color = "#abb2b9", size = cty_line) +
  geom_sf(data = USstates, fill = NA) +
  scale_fill_brewer(palette = "OrRd", guide = guide_legend(reverse = TRUE)) +
  coord_sf(crs = st_crs(5070)) +
  facet_grid(FIRE_YEAR ~ ., switch = "y") +
  theme_void() +
  labs(fill = "# of fires")
map_7
```
\newpage

Adjustments to facet label orientation and text size

```{r map improve map 8, message=FALSE, warning=FALSE}
text_size <- 15

map_8 <- ggplot(UScounties, aes(fill = fire_sum_total_cat)) +
  geom_sf(color = "#abb2b9", size = cty_line) +
  geom_sf(data = USstates, fill = NA) +
  scale_fill_brewer(palette = "OrRd", guide = guide_legend(reverse = TRUE)) +
  coord_sf(crs = st_crs(5070)) +
  facet_grid(FIRE_YEAR ~ ., switch = "y") +
  theme_void() +
  theme(
    text = element_text(size = text_size),
    strip.text.y.left = element_text(angle = 0) 
        ) +
  labs(fill = "# of fires")
map_8
```
\newpage

## Map example 2 - Socioeconomic status (SES)

```{r map SES example, message=FALSE, warning=FALSE, echo=FALSE, results="hide"}
###
#### Data prep
###

SES_df <- readRDS(paste0(external_data_dir, "SES_ex\\SES_ex.rds"))

# Factor analysis
SES_mo_var <- SES_df[,-c("FIPS")]
# parallel <- fa.parallel(SES_mo_var, fm = 'minres', fa = 'fa') 
fit <- factanal(SES_mo_var, 2, rotation = "varimax", scores = "regression")
Pov_factor <- data.table(Pov_factor = fit$scores[,1])
FIPS_SES <- cbind(SES_df[,c("FIPS")], Pov_factor)

# Quantile groups of SES factor
FIPS_SES$Pov_factor_5 <- cut(FIPS_SES$Pov_factor, breaks = quantile(FIPS_SES$Pov_factor, probs = seq(0, 1, length.out = 6)),
                             labels = c(as.character(seq(1:5))))
FIPS_SES$Pov_factor_10 <- cut(FIPS_SES$Pov_factor, breaks = quantile(FIPS_SES$Pov_factor, probs = seq(0, 1, length.out = 11)),
                             labels = c(as.character(seq(1:10))))
# table(FIPS_SES$Pov_factor_5)
# table(FIPS_SES$Pov_factor_10)

### State and county shapefiles (US)
census_shapfile_dir <- paste0(external_data_dir, "Shapefiles\\Census_2010\\")
USstates <- st_read(paste0(census_shapfile_dir, "States2010_outline\\gz_2010_us_040_00_5m\\gz_2010_us_040_00_5m.shp"))
UScounties <- st_read(paste0(census_shapfile_dir, "Counties2010_outline\\gz_2010_us_050_00_5m\\gz_2010_us_050_00_5m.shp"))
UScounties$FIPS <- paste0(UScounties$STATE, UScounties$COUNTY)

### Join aggregated fire data to county shapefile
UScounties <- UScounties %>%
  left_join(FIPS_SES, by = c("FIPS"))

### Subset to continuous US (exclude HI, AK, PR)
exclude_HI_AK_PR <- c("02","15","72")
USstates <- USstates %>% filter( !(STATE %in% exclude_HI_AK_PR) )
UScounties <- UScounties %>% filter( !(STATE %in% exclude_HI_AK_PR))
```

Default map for 5 SES groups

```{r map SES 5 group ex 1, message=FALSE, warning=FALSE}
cty_line <- 0.2

map_1 <- ggplot(UScounties, aes(fill = Pov_factor_5)) +
  geom_sf(size = cty_line) +
  geom_sf(data = USstates, fill = NA) +
  coord_sf(crs = st_crs(5070)) +
  theme_void() +
  labs(fill = "SES level 5 groups")
map_1
```
\newpage

Color palette - one continuous color palate

```{r map SES 5 group ex 2, message=FALSE, warning=FALSE}
map_2 <- ggplot(UScounties, aes(fill = Pov_factor_5)) +
  geom_sf(size = cty_line) +
  geom_sf(data = USstates, size = 0.5, fill = NA) +
  scale_fill_brewer(palette = "Greens", na.value = "#e8e6e4") +
  coord_sf(crs = st_crs(5070)) +
  theme_void() +
  labs(fill = "SES level 5 groups")
map_2
```
\newpage

Color palette - discrete categories

```{r map SES 5 group ex 3, message=FALSE, warning=FALSE}
map_3 <- ggplot(UScounties, aes(fill = Pov_factor_5)) +
  geom_sf(size = cty_line) +
  geom_sf(data = USstates, size = 0.5, fill = NA) +
  scale_fill_brewer(palette = "Dark2", na.value = "#e8e6e4") +
  coord_sf(crs = st_crs(5070)) +
  theme_void() +
  labs(fill = "SES level 5 groups")
map_3
```
\newpage

Color palette - discrete categories

```{r map SES 5 group ex 4, message=FALSE, warning=FALSE}
colors_5 <- c("#EE7600", "#82d6b4", "#3da078", "#057f4e", "#483D8B")

map_4 <- ggplot(UScounties, aes(fill = Pov_factor_5)) +
  geom_sf(color = "grey62", size = cty_line) +
  geom_sf(data = USstates, size = 0.5, fill = NA) +
  scale_fill_manual(values = colors_5, na.value = "#e8e6e4") +
  # scale_fill_manual(values = colors_5, na.value = "grey90") +
  coord_sf(crs = st_crs(5070)) +
  theme_void() +
  labs(fill = "SES level 5 groups")
map_4
```
\newpage

Default map for 10 SES groups

```{r map SES 10 group ex 1, message=FALSE, warning=FALSE}
map_10_1 <- ggplot(UScounties, aes(fill = Pov_factor_10)) +
  geom_sf(size = cty_line) +
  geom_sf(data = USstates, fill = NA) +
  scale_fill_brewer(palette = "Greens", na.value = "#e8e6e4") +
  coord_sf(crs = st_crs(5070)) +
  theme_void() +
  labs(fill = "SES level 10 groups")
map_10_1
```
\newpage

Create levels for map w/ 10 SES groups
```{r map SES 10 group ex 2, message=FALSE, warning=FALSE}
colourCount <- 10
colors_use <- colorRampPalette(brewer.pal(8, "Greens"))(colourCount)

map_10_2 <- ggplot(UScounties, aes(fill = Pov_factor_10)) +
  geom_sf(size = cty_line) +
  geom_sf(data = USstates, fill = NA) +
  scale_fill_manual(values = colors_use, na.value = "#e8e6e4") +
  coord_sf(crs = st_crs(5070)) +
  theme_void() +
  labs(fill = "SES level 10 groups")
map_10_2
```
\newpage




